Power to the Stimuli: Not the Effect

### 2023-03-29 manuscript_draft.Rmd
so much all the updates

bibliography      : ["references.bib"]

@@

Changes in publication practices and research design have also shown a new wrinkle to providing a sample size plan for a research study. While statistics courses often suggest that a specific research design leads to a specific statistical test, multiple Many Analysts papers have shown that - given the same data and hypothesis - researchers can come up with multiple ways to analyze the data [@silberzahn2018many, @coretta]. Research projects often have multiple testable hypothesis, however, it is unclear which hypothesis or test the sample size planning should be performed on. Further, an entire set of common research publications may not even have a hypothesis to examine within their project, as they are simply providing a large, quality dataset for future reuse [i.e., stimuli database creation [@buchanan2019]. The increased ability to compute complex statistical analyses, such as multilevel modeling, has pushed researchers with repeated measures designs to abandon creating person-level averages just to be able to use a traditional ANOVA [@brysbaert2018]. These analyses have also made it clear that we should be careful to assume that all items in a research study have the "effect", as there is often variability in their impact on the outcomes of the study.

@@

In this approach, a researcher could use sequential testing to estimate their parameter of interest after each participant to determine if they have achieved their expected width of the confidence interval around that parameter. In this procedure, one should consider setting a minimum sample size (i.e., based on known data collection ability), use the confidence interval width as a stopping rule (i.e., stop data collect when the CI is narrow), and use the estimated sample size from the AIPE calculations as a potential maximum sample size. By defining each of these components, a research could ensure a feasible minimum sample size, a way to cease data collection when goals have been met, and a stopping rule to ensure an actual end to data collection. Given pilot or previously collected data, we should be able to leverage the ideas behind AIPE, paired with simulation and bootstrapping, to estimate the minimum and maximum proposed sample sizes and stopping rules for repeated measures studies with expected variabillity in parameter estimates for items.

@@

Using ideas from newer power methods, we suggest the following procedure to determine a sample size for each item:

@@

In this section, we provide two examples of the suggested procedure. The first example includes concreteness ratings from Brysbaert et al. [CITE, 2014]. Instructions given to participants denoted the difference between concrete (i.e., "refers to something that exists in reality") and abstract (i.e., "something you cannot experience directly through your senses or actions") terms. Participants were then asked to rate concreteness of terms using a 1 (abstract) to 5 (concrete) scale. This data represents a small scale dataset that could be used as pilot data for a study using concrete word ratings. The data is available at OSF LINK [CITE]. The second dataset includes a large scale dataset with response latencies, the English Lexicon Project [CITE, Balota et al.]. The English Lexicon Project consists of lexical decision response latencies for English words. In a lexical decision task, participants simply select "word" for real words (e.g., *dog*) and "nonword" for pseudowords (e.g., *wug*). The trial level data is available here [<https://elexicon.wustl.edu/>, CITE]. Critically, in each of these examples, the individual trial level data for each item is available to simulate and calculate standard errors on. Data that has been summarized could potentially be used, as long as the original standard deviations for each item were present. From the mean and standard deviation for each item, a simulated pilot dataset could be generated for estimating new sample sizes. All code to estimate sample sizes is provided on our OSF page.

@@

While the examples in this manuscript are traditionally cognitive linguistics focused, any research using repeated items can benefit from newer sampling techniques. Therefore, we provide XX example vignettes on our OSF page/GitHub site for this manuscript across a range of examples of data types provided by the authors of this manuscript. Examples include ... ADD HERE AFTER DONE WITH VIGNETTES.

@@

To be added once people say this idea doesn't suck.

>>>>

@@

Changes in publication practices and research design have also shown a new wrinkle to providing a sample size plan for a research study. While statistics courses often suggest that a specific research design leads to a specific statistical test, multiple Many Analysts papers have shown that - given the same data and hypothesis - researchers can come up with multiple ways to analyze the data [@silberzahn2018many, @coretta]. Research projects often have multiple testable hypothesis, however, it is unclear which hypothesis or test the sample size planning should be performed on. Further, an entire set of common research publications may not even have a hypothesis to examine within their project, as they are simply providing a large, quality dataset for future reuse [i.e., stimuli database creation [@buchanan2019]. The increased ability to compute complex statistical analyses, such as multilevel modeling, has pushed researchers with repeated measures designs to abandon creating person-level averages just to be able to use a traditional ANOVA [@brysbaert2018]. These analyses have also made it clear that we should be careful to assume that all items in a research study have the same "effect", as there is often variability in their impact on the outcomes of the study.

@@

In this approach, a researcher could use sequential testing to estimate their parameter of interest after each participant to determine if they have achieved their expected width of the confidence interval around that parameter. One would set a minimum sample size (i.e., based on known data collection ability), use the confidence interval width as a stopping rule (i.e., stop data collect when the CI is narrow), and use the estimated sample size from the AIPE calculations as a potential maximum sample size. By defining each of these components, a research could ensure a feasible minimum sample size, a way to cease data collection when goals have been met, and a stopping rule to ensure an actual end to data collection. Given pilot or previously collected data, we should be able to leverage the ideas behind AIPE, paired with simulation and bootstrapping, to estimate the minimum and maximum proposed sample sizes and stopping rules for repeated measures studies with expected variabillity in parameter estimates for items.

@@

Using these ideas, we suggest the following procedure to determine a sample size for each item:

@@

In this section, we provide two examples of the suggested procedure. The first example includes concreteness ratings from @brysbaert2014. Instructions given to participants denoted the difference between concrete (i.e., "refers to something that exists in reality") and abstract (i.e., "something you cannot experience directly through your senses or actions") terms. Participants were then asked to rate concreteness of terms using a 1 (abstract) to 5 (concrete) scale. This data represents a small scale dataset that could be used as pilot data for a study using concrete word ratings. The data is available at <https://osf.io/qpmf4/>. The second dataset includes a large scale dataset with response latencies, the English Lexicon Project [@balota2007]. The English Lexicon Project consists of lexical decision response latencies for English words. In a lexical decision task, participants simply select "word" for real words (e.g., *dog*) and "nonword" for pseudowords (e.g., *wug*). The trial level data is available here [<https://elexicon.wustl.edu/>]. Critically, in each of these examples, the individual trial level data for each item is available to simulate and calculate standard errors on. Data that has been summarized could potentially be used, as long as the original standard deviations for each item were present. From the mean and standard deviation for each item, a simulated pilot dataset could be generated for estimating new sample sizes. All code to estimate sample sizes is provided on our OSF page.

@@

While the examples in this manuscript are traditionally cognitive linguistics focused, any research using repeated items can benefit from newer sampling techniques. Therefore, we provide XX example vignettes and code examples on our OSF page/GitHub site for this manuscript across a range of examples of data types provided by the authors of this manuscript. Examples include psycholinguistics, social psychology, COVID related data, and cognitive psychology.

@@

In this manuscript, we demonstrated a method using AIPE and simulation/bootstrapping to estimate a minimum and maximum sample size along with a rule for stopping data collection based on narrow confidence intervals on the parameter of interest. We believe this procedure is specifically useful for studies with multiple items that intend on using item level focused analyses; however, the utility of measuring each item well can extend to many analysis choices. By focusing on gathering quality data, we can suggest that the data is useful, regardless of outcome of any hypothesis test.

One limitation of these methods would be using datasets with very large numbers of items to simulate what might happen within one study. For example, the English Lexicon Project includes thousands of items, and by the time we would simulate for all of those, it would likely suggest needing thousands of participants for *most items* to reach criterion. Alternatively, as the number of items increases, you also could potentially see very small estimates for sample size due to the correction factor (as with large numbers of items, you could find many items with standard errors below the 50% decile). Therefore, it would be beneficial to consider only simulating with what a participant would reasonably complete in a study. On the other side, small numbers of repeated items usually result in higher sample sizes proposed from the original pilot data. This result occurs because the smaller number of items means more samples for nearly all to reach the cutoff criteria. These results are not too different than what we might expect for a power analysis using a multilevel model - larger number of items tends to decrease necessary sample size, while smaller numbers of items tend to increase sample size.

Second, these methods do not ensure the normal interpretation of power, wherein you know would find a specific effect for a specific test, $\alpha$, and so on. As discussed in the introduction, there is not necessarily a one-to-one mapping of hypothesis to analysis, and many of the guesses within a traditional power analysis are just that - best guesses for various parameters. These methods could be used together to strengthen our understanding of sample size necessary for both a hypothesis test and well tuned estimation.
